\subsection{User-defined functions}\label{user-defined-function-section}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{User-defined distributions}\label{user-defined-distribution-section}

Probability distributions are implemented in Java.  Distribution classes should implement the interface \verb|blog.distrib.CondProbDistrib|. 
By default, the \bl engine will look up distribution classes in the package \verb|blog.distrib|. In addition, it will look up distribution classes under the default empty package. 

Below is one example of a uniform distribution on Integers.
\begin{minted}{java}
package blog.distrib;

import blog.common.Util;

public class UniformInt implements CondProbDistrib {

  /**
   * set parameters for UniformInt
   * distribution
   * 
   * @param params
   *          An array of [Integer, Integer]
   *          <ul>
   *          <li>params[0]: <code>lower</code> (Integer)</li>
   *          <li>params[1]: <code>upper</code> (Integer)</li>
   *          </ul>
   * 
   * @see blog.distrib.CondProbDistrib#setParams(java.util.List)
   */
  @Override
  public void setParams(Object[] params) {
    if (params.length != 2) {
      throw new IllegalArgumentException("expected two parameters");
    }
    setParams((Integer) params[0], (Integer) params[1]);
  }

  /**
   * For a non-null value of method parameter lower, sets the
   * distribution parameter <code>lower</code> to method parameter lower.
   * Similarly for <code>upper</code>. Then checks to see if assignment of
   * parameters is legal. In other words, an assignment of parameters is legal
   * if <code>lower <= upper</code>.
   * 
   * @param lower
   *          parameter <code>lower</code>
   * @param upper
   *          parameter <code>upper</code>
   */
  public void setParams(Integer lower, Integer upper) {
    if (lower != null) {
      this.lower = lower;
      this.hasLower = true;
    }
    if (upper != null) {
      this.upper = upper;
      this.hasUpper = true;
    }
    if (this.hasLower && this.hasUpper) {
      if (this.lower > this.upper) {
        throw new IllegalArgumentException(
            "UniformInt distribution requires that lower <= upper");
      }
      this.prob = 1.0 / (this.upper - this.lower + 1);
      this.logProb = Math.log(this.prob);
    }
  }

  private void checkHasParams() {
    if (!this.hasLower) {
      throw new IllegalArgumentException("parameter lower not provided");
    }
    if (!this.hasUpper) {
      throw new IllegalArgumentException("parameter upper not provided");
    }
  }

  /*
   * (non-Javadoc)
   * 
   * @see blog.distrib.CondProbDistrib#getProb(java.lang.Object)
   */
  @Override
  public double getProb(Object value) {
    return getProb(((Integer) value).intValue());
  }

  /**
   * Returns the probability of the UniformInt distribution having outcome
   * <code>value</code>.
   */
  public double getProb(int value) {
    checkHasParams();
    return (value >= lower) && (value <= upper) ? prob : 0;
  }

  /*
   * (non-Javadoc)
   * 
   * @see blog.distrib.CondProbDistrib#getLogProb(java.lang.Object)
   */
  @Override
  public double getLogProb(Object value) {
    return getLogProb(((Integer) value).intValue());
  }

  /**
   * Returns the log probability of the UniformInt distribution having outcome
   * <code>value</code>.
   */
  public double getLogProb(int value) {
    checkHasParams();
    return ((value >= lower) && (value <= upper)) ? logProb
        : Double.NEGATIVE_INFINITY;
  }

  /*
   * (non-Javadoc)
   * 
   * @see blog.distrib.CondProbDistrib#sampleVal()
   */
  @Override
  public Object sampleVal() {
    return sample_value();
  }

  public int sample_value() {
    checkHasParams();
    return lower + Util.randInt(upper - lower + 1);
  }

  @Override
  public String toString() {
    return getClass().getName();
  }

  /** Parameter <code>lower</code>. */
  private int lower;
  /** Flag indicating whether <code>lower</code> has been set. */
  private boolean hasLower;
  /** Parameter <code>upper</code>. */
  private int upper;
  /** Flag indicating whether <code>upper</code> has been set. */
  private boolean hasUpper;
  /**
   * The probability of an outcome between <code>lower</code> and
   * <code>upper</code> inclusive.
   */
  private double prob;
  /**
   * The log probability of an outcome between <code>lower</code> and
   * <code>upper</code> inclusive.
   */
  private double logProb;
}
\end{minted}


\section{Comprehensive examples}\label{example-section}

\subsection{Burglary}
The following example was due initially to Judea Pearl:
the alarm in a house goes off in response to a burglary or an earthquake,
but is somewhat unreliable. We might write the following model:
\begin{blogcode}
type House;

distinct House Maryhouse, Johnhouse, Cathyhouse, Rogerhouse;

random Boolean Burglary(House h) ~ BooleanDistrib(0.003);
random Boolean Earthquake ~ BooleanDistrib(0.002);
random Boolean Alarm(House h) ~ 
  case [Burglary(h), Earthquake] in {
    [false, false] -> BooleanDistrib(0.01),
    [false, true]  -> BooleanDistrib(0.40),
    [true, false]  -> BooleanDistrib(0.80),
    [true, true ]  -> BooleanDistrib(0.90)
  };

obs Alarm(Maryhouse) = true;
obs Alarm(Johnhouse) = true;
obs Alarm(Cathyhouse) = true;
obs Alarm(Rogerhouse) = false;

query Earthquake;
\end{blogcode}

\bl inference engine will produce the following likelihood of earthquake.
\begin{verbatim}
Number of samples: 10000
Distribution of values for Earthquake
	true	0.967140181036552
	false	0.03285981896344725
\end{verbatim}


\subsection{A hidden Markov model for genetic sequences}
\begin{example}[Hidden Markov models]
The following represents a hidden Markov model for genetic sequences with four states and four output symbols. The state at each time step transitions to another with respect to a conditional distribution specified by a TabularCPD. 
Each state at each time step emits an observation with respect to another CPD. After making a few observations, we can query the states for each time step.
\end{example}

\begin{blogcode}
type State;
distinct State A, C, G, T;

type Output;
distinct Output ResultA, ResultC, ResultG, ResultT;

random State S(Timestep t) ~
  if t == @0 then 
    Categorical({A -> 0.3, C -> 0.2, G -> 0.1, T -> 0.4})
  else case S(prev(t)) in {
    A -> Categorical({A -> 0.1, C -> 0.3, G -> 0.3, T -> 0.3}),
    C -> Categorical({A -> 0.3, C -> 0.1, G -> 0.3, T -> 0.3}),
    G -> Categorical({A -> 0.3, C -> 0.3, G -> 0.1, T -> 0.3}),
    T -> Categorical({A -> 0.3, C -> 0.3, G -> 0.3, T -> 0.1})
  };

random Output O(Timestep t) ~ 
  case S(t) in {
    A -> Categorical({
      ResultA -> 0.85, ResultC -> 0.05, 
      ResultG -> 0.05, ResultT -> 0.05}),
    C -> Categorical({
      ResultA -> 0.05, ResultC -> 0.85, 
      ResultG -> 0.05, ResultT -> 0.05}),
    G -> Categorical({
      ResultA -> 0.05, ResultC -> 0.05, 
      ResultG -> 0.85, ResultT -> 0.05}),
    T -> Categorical({
      ResultA -> 0.05, ResultC -> 0.05, 
      ResultG -> 0.05, ResultT -> 0.85})
  };

/* Evidence for the Hidden Markov Model.
 */

obs O(@0) = ResultC;
obs O(@1) = ResultA;
obs O(@2) = ResultA;
obs O(@3) = ResultA;
obs O(@4) = ResultG;

/* Queries for the Hiddem Markov Model, given the evidence.  
 * Note that we can query S(5) even though our observations only 
 * went up to time 4.
 */

query S(@0);
query S(@1);
query S(@2);
query S(@3);
query S(@4);
query S(@5);
\end{blogcode}

\bl will generate the following results using the particle filtering algorithm. 
\begin{verbatim}
Distribution of values for S(@0)
	C	0.8128524436090175
	T	0.09473684210526356
	A	0.06905545112781902
	G	0.023355263157894345
Distribution of values for S(@1)
	A	0.8700181159420364
	G	0.05365942028985556
	T	0.05143115942029019
	C	0.024891304347826236
Distribution of values for S(@2)
	A	0.7075761624799592
	C	0.09974612506680965
	G	0.0965058792089793
	T	0.0961718332442546
Distribution of values for S(@3)
	A	0.7633727477477481
	C	0.08085585585585582
	G	0.07908220720720713
	T	0.07668918918918921
Distribution of values for S(@4)
	G	0.8739530096536214
	C	0.050912123793299964
	T	0.05014906303236865
	A	0.02498580352072714
Distribution of values for S(@5)
	A	0.2989778534923363
	C	0.2949673480976736
	T	0.2767248722316938
	G	0.12932992617830938
\end{verbatim}

\appendix

\phantomsection
\addcontentsline{toc}{section}{\appendixtocname}


\section{A Brief  History of BLOG}\label{history-appendix}
Bayesian Logic (BLOG) was first developed by Brian Milch in 2005. Since then, major contribution is from various members of Professor Stuart Russell's research group at University of California Berkeley. 

The initial syntax and semantics of \bl was described in 

\begin{quote}
Brian Milch, Bhaskara Marthi, Stuart Russell, David Sontag, Daniel L. Ong, and Andrey Kolobov (2005) "BLOG: Probabilistic Models with Unknown Objects". Proc. 19th International Joint Conference on Artificial Intelligence (IJCAI): 1352-1359.
\end{quote}

Dr. Rodrigo Braz introduced Timestep into BLOG. Milch and Braz released \bl 0.3 in 2008.

Dr. Lei Li has been leading the development of the language and \bl inference system since 2011. Since then, the language sees major changes, as well as the inference engine. New algorithms are introduced. However, the semantics of \bl remain the same as the original. 

\section{\bl Grammar Definition}\label{bnf-appendix}

\begin{verbatim}
program ::= opt_statement_lst ;

opt_statement_lst ::= 
    /* EMPTY */
  | statement_lst ;

statement_lst ::= 
    statement SEMI statement_lst
  | statement statement_lst
  | statement SEMI 
  | statement ;

statement ::= 
    declaration_stmt  
  | evidence_stmt  
  | query_stmt ;

declaration_stmt ::=
    type_decl 
  | fixed_func_decl 
  | rand_func_decl 
  | origin_func_decl  
  | number_stmt  
  | distinct_decl  
  | parameter_decl  
  | distribution_decl ;

type_decl ::= TYPE ID ;

type ::= 
    refer_name
  | list_type 
  | array_type  
  | map_type ;

list_type ::= LIST LT refer_name GT ;

array_type_or_sub ::= refer_name LBRACKET ;

array_type ::= 
    array_type_or_sub RBRACKET
  | array_type  LBRACKET RBRACKET ;

map_type ::= MAP LT type COMMA type GT ;

opt_parenthesized_type_var_lst ::= 
    /* EMPTY */
  | parenthesized_type_var_lst
  | type_var_lst ;

parenthesized_type_var_lst ::= 
    LPAREN RPAREN
  | LPAREN type_var_lst RPAREN ;

extra_commas ::= 
    COMMA COMMA
  | extra_commas COMMA ;

type_var_lst ::= 
    type ID COMMA type_var_lst
  | type ID
  | type ID extra_commas type_var_lst
  | type ID type_var_lst
  | type COMMA type_var_lst ;

fixed_func_decl ::=
    FIXED type_type ID opt_parenthesized_type_var_lst
    EQ expression ;

rand_func_decl ::=
    RANDOM type_type ID opt_parenthesized_type_var_lst
    dependency_statement_body ;

number_stmt ::=
    NUMSIGN refer_name opt_parenthesized_origin_var_list
  dependency_statement_body
  | NUMSIGN opt_parenthesized_origin_var_list
  dependency_statement_body ;

opt_parenthesized_origin_var_list ::= 
    /* Empty */
  | LPAREN origin_var_list RPAREN ;

origin_var_list ::= 
    ID EQ ID COMMA origin_var_list
  | ID EQ ID extra_commas origin_var_list
  | ID EQ COMMA origin_var_list
  | ID EQ ID origin_var_list
  | ID EQ ID
  | ID ID ;

origin_func_decl ::=
  ORIGIN type_type ID LPAREN type_type RPAREN 
 | ORIGIN type_type LPAREN type_type RPAREN 
 | ORIGIN type_type ID LPAREN type_type 
 | ORIGIN type_type ID type_type RPAREN ;

distinct_decl ::=
    DISTINCT refer_name id_or_subid_list ;

id_or_subid_list ::= 
   id_or_subid 
 | id_or_subid COMMA id_or_subid_list
 | id_or_subid id_or_subid_list
 | id_or_subid extra_commas id_or_subid_list ;

id_or_subid ::= 
   ID
 | ID LBRACKET INT_LITERAL RBRACKET ;

distribution_decl ::=
    DISTRIBUTION ID EQ refer_name
    LPAREN opt_expression_list RPAREN ;

refer_name ::= 
    ID 
  | ID DOT refer_name ;

dependency_statement_body ::= DISTRIB expression ;

parameter_decl ::= 
   PARAM type ID 
 | PARAM type ID COLON expression ;

expression ::= 
    operation_expr  
  | literal  
  | function_call 
  | list_expr 
  | map_construct_expression 
  | quantified_formula 
  | set_expr 
  | number_expr 
  | if_expr 
  | case_expr ;

literal ::= 
   STRING_LITERAL
 | CHAR_LITERAL
 | INT_LITERAL
 | DOUBLE_LITERAL
 | BOOLEAN_LITERAL
 | NULL ;

operation_expr ::= 
  expression PLUS expression
 | expression MINUS expression
 | expression MULT expression
 | expression DIV expression
 | expression MOD expression
 | expression POWER expression
 | expression LT expression
 | expression GT expression
 | expression LEQ expression
 | expression GEQ expression
 | expression EQEQ expression
 | expression NEQ expression
 | expression AND expression
 | expression OR expression
 | expression DOUBLERIGHTARROW expression
 | expression LBRACKET expression RBRACKET
 | unary_operation_expr ;

unary_operation_expr ::=
   MINUS expression %prec UMINUS
 | NOT expression
 | AT expression
 | LPAREN expression RPAREN ;  

quantified_formula ::= 
   FORALL type ID expression
 | EXISTS type ID expression ;

function_call ::= 
   refer_name LPAREN opt_expression_list RPAREN
 | refer_name ;

if_expr ::= 
   IF expression THEN expression ELSE expression 
 | IF expression THEN expression ;

case_expr ::= CASE expression IN map_construct_expression ;

opt_expression_list ::= 
    expression_list 
  | /*EMPTY*/ ;

expression_list ::= 
   expression COMMA expression_list
 | expression
 | expression extra_commas expression_list ;

semi_colon_separated_expression_list ::=
   semi_ending_expression_list semi_colon_separated_expression_list
 | semi_ending_expression_list expression_list ;

semi_ending_expression_list ::=
   expression_list SEMI
 | semi_ending_expression_list SEMI ;

map_construct_expression ::= LBRACE expression_pair_list RBRACE ;

expression_pair_list ::=
   expression RIGHTARROW expression COMMA expression_pair_list
 | expression RIGHTARROW expression ;
  
number_expr ::= 
   NUMSIGN set_expr
 | NUMSIGN type ;

list_expr ::= 
   LBRACKET opt_expression_list RBRACKET
 | LBRACKET semi_colon_separated_expression_list RBRACKET
 | LBRACKET comprehension_expr RBRACKET ;

set_expr ::= 
    explicit_set 
  | tuple_set ;

explicit_set ::= LBRACE opt_expression_list RBRACE ;

comprehension_expr ::= expression_list FOR type_var_lst opt_colon_expr ;

opt_colon_expr ::= 
    /* EMPTY */
  | COLON expression ;

tuple_set ::= LBRACE comprehension_expr RBRACE ;

evidence_stmt ::= OBS evidence ;

evidence ::= value_evidence ;

value_evidence ::= expression EQ expression ;

query_stmt ::= QUERY expression ;
\end{verbatim}

